---
title: "Introduction to getDTeval"
output: rmarkdown::html_vignette
author: 'Authors: David Shilane, Caffrey Lee, Zoe Huang, Anderson Nelson'
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Introduction to getDTeval}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r library, include=FALSE}
library(formulaic)
library(data.table)
library(stats)
library(microbenchmark)
library(dplyr)
library(knitr)
library(getDTeval)
dat = formulaic::snack.dat
```

```{r constant,echo=FALSE}
age.name = "Age"
mean.age.name = sprintf("Mean %s", age.name)
awareness.name = "Awareness"
gender.name = "Gender"
region.name = "Region"
```

# Introduction

Using the `get()` and `eval()` functions allows for more programmatic coding designs that enable greater flexibility and more dynamic computations. However, in data.table statements, `get()` and `eval()` reduce the efficiency of the method by performing work prior to data.table's optimized computations. **getDTeval** is useful in translateing `get()` and `eval()` statements more efficiently for improved runtime performance.

**Note:** getDTeval would not perform any conversions of data types. When the statement is for data table, to make sure that the data type of the `dat` would always be data table, we would run `dat = as.data.table(dat)` as the first step in all examples.

# Functions and Applications

**getDTeval** package has 2 main functions. The main purpose of developing the package is to translates `get()` and `eval()` statements more efficiently, which allows a user to both incorporate programmatic designs and while utilizing data.table's efficient processing routines.

 - **benchmark.getDTeval** performs a benchmarking experiment for data.table coding statements that use `get()` or `eval()` for programmatic designs.  The a) original statement is compared to b) passing the original statement through getDTeval and also to c) an optimized coding statement.  The results can demonstrate the overall improvement of using the coding translations offered by `getDTeval::getDTeval()`.

 - **getDTeval** offers a method of translating coding statements into more efficient versions for improved runtime performance.

**1. Returning the translated coding statement:**

```{r getDTeval1}
dat = as.data.table(dat)
the.statement <- "dat[get(gender.name) == 'Female', mean(get(age.name)), keyby = region.name]"
getDTeval(the.statement = the.statement, return.as = "code")
``` 

**2. Returning the calculation result:**

```{r getDTeval2}
dat = as.data.table(dat)
getDTeval(the.statement = the.statement, return.as = "result")
``` 

**3. Returning the a list of the calculation result and the code:**

```{r getDTeval3}
dat = as.data.table(dat)
getDTeval(the.statement = the.statement, return.as = "all")
``` 

# Benefits

There are some major applications to the getDTeval package:

**1). Combining programmatic coding designs with data.table's efficiency. Better utilizing `get()` and `eval()` without the trade-offs in performance.**

This is a benchmarking experiment on the snack.dat from formulaic package. The result shows the reduction in running time using getDTeval over the original statement.

```{r benchmark.getDTeval_example}
sample.dat <- dat[sample(x = 1:.N, size = 10^6, replace = TRUE)]
sample.dat = as.data.table(sample.dat)
the.statement <- "sample.dat[get(age.name) > 65, .(mean_awareness = mean(get(awareness.name))), keyby = c(eval(gender.name), region.name)]"
benchmark.getDTeval(the.statement = the.statement)
```

**2). Expanding on the use of `eval()` in data.table's calculations.** 

In data.table's cacluations, `eval()` within a `.()` calculation in the j step is not allowed. For example, if we try to run `dat[, .(eval(mean.age.name) = mean(get(age.name)))]`, then the error message would be `Error: unexpected '=' in "dat[, .(eval(mean.age.name) ="`

```{r getDTeval_example1_with_error, error = TRUE, purl = FALSE}
dat[, .(eval(mean.age.name) = mean(get(age.name)))]
```

However, using **getDTeval** allows this to work:

```{r getDTeval_example1, eval=FALSE}
dat = as.data.table(dat)
getDTeval(the.statement = "dat[, .(eval(mean.age.name) = mean(get(age.name)))]")
```

**3). Expanding on the use of `eval()` in dplyr code.**

`eval()` steps in dplyr code is not allowed. For example, here we have a data frame:

```{r setup dataframe}
df = as.data.frame(dat)
```

If we try to run `df %>% summarize(eval(mean.age.name) = mean(get(age.name)))`, then the error message would be `Error: unexpected '=' in "df %>% summarize(eval(mean.age.name) ="`

```{r getDTeval_example2_with_error, error = TRUE, purl = FALSE}
df %>% summarize(eval(mean.age.name) = mean(get(age.name)))
```

However, using **getDTeval** allows this to work.

```{r getDTeval_example2}
getDTeval(the.statement = "df %>% summarize(eval(mean.age.name) = mean(get(age.name)))")
```
